Context-dependent consequences of including lagged effects in demographic models

### 2023-07-20 docs/ipm_comparison_ms.tex
adding quarto version of ms

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyfoot{}
\fancyhead[R]{p. \thepage}
\fancyhead[L]{Draft - 18 July 2023}
\usepackage{sectsty} \sectionfont{\centering}
\DeclareUnicodeCharacter{U+2206}{$\delta$}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Context-dependent consequences of including lagged effects in demographic models},
  pdfkeywords={integral projection models; environmental stochasticity; lagged effects},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Context-dependent consequences of including lagged effects in demographic models}
\author{true \and true}
\date{18 July, 2023}

\begin{document}
\maketitle

Keywords: integral projection models; environmental stochasticity; lagged effects

Running title: Lagged Effects IPMs

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Demographic models are widely used for all kinds of stuff
\item
  It has long been recognized that there is the potential for lagged effects.
  Now there's evidence for that (Evers et al. 2021; Scott et al. 2022).
\item
  Lagged effects could be different in different habitat types (and this could be a big reason for differences between habitats)
\end{enumerate}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{demographic-data}{%
\subsection{Demographic Data}\label{demographic-data}}

\hypertarget{ipms}{%
\subsection{IPMs}\label{ipms}}

@@

We built three classes of IPMs for comparison which each required different functional forms of their underlying vital rates models.
The simplest IPM was a general, density-independent, deterministic IPM with four sub-kernels: growth and survival (\(P\), (\textbf{eq:P?})), fecundity (\(F\), i.e.~production of new seedlings, (\textbf{eq:F?})), probability of staying a seedling (always 0), and recruitment (\(R\), i.e.~seedling survival and establishment, (\textbf{eq:R?})) ((\textbf{eq:mature?}; \textbf{eq:sdlg?}), Figure \ref{fig:lifecycle}).
The probability of staying a seedling, was always equal to zero, since our definition of seedlings was first year plants only.

@@

To build the general, density-independent, stochastic, kernel-resampled IPMs, we included environmental stochasticity in all vital rate models built using the long term demographic datset by adding a random effect of year (Figure \ref{fig:lifecycle}).
The random effect of year was included using a factor--smooth interaction which allowed the relationship between plant size and vital rates to vary in functional form among transition years.
The kernel-resampling approach is to generate kernels corresponding to each transition year in the demographic dataset using the random smooths for year, and to iterate the IPM by drawing from these randomly.
This is equivalent to the matrix selection approach for matrix population models described by Caswell (2001).

For the third method, we modeled the impacts of drought on vital rates explicitly and created general, density-independent, stochastic, parameter-resampled IPMs (\emph{sensu} Metcalf et al. (2015)).
We calculated the standardized precipitation evapotranspiraton index (SPEI) for our site using a published gridded dataset based on ground measurements (Xavier et al. 2016) as described in Scott et al. (2022).
For all vital rate models fit using the long term demographic dataset, we modeled delayed effects of SPEI using distributed lag non-linear models with a maximum lag of 36 months(Scott et al. 2022) (Figure \ref{fig:lifecycle}).
To iterate these parameter-resampled IPMs, a random sequence of SPEI values was created by sampling years of the observed monthly SPEI data.
Then, 36 month lags are calculated for each year starting in February (the month of the demographic census).
These values are then used to predict fitted values from the vital rates models, generating different kernels at each iteration of the IPM.
With this method, the kernels of successive iterations are not entirely independent because the SPEI values used in calculating vital rates include values used in the previous two iterations, but they are ergodic.

@@

\hypertarget{results}{%
\section{Results}\label{results}}

For all vital rates estimated using the long term demographic dataset, the DLNM model fit the best (dAIC = 0) followed by the model with a random effect of year, followed by the deterministic model (Table \ref{tab:aic}).

Population growth rates were consistently higher in continuous forest compared to forest fragments across IPM types (Table \ref{tab:lambdas}).

The time to iterate the DLNM models is much higher than than deterministic and kernel-resampled.
The greater use of computational resources is likely a result of \texttt{predict()} being much slower for GAMs with 2D smooths because the number of knots is much higher compared to the GAMs used for the vital rates models in the determinsitic and kernel-resampled IPMs.

\begin{verbatim}
#> # A tibble: 3 x 2
#>   IPM   mean_time_min
#>   <chr>         <dbl>
#> 1 det            0.02
#> 2 dlnm          87.1 
#> 3 stoch          0.07
\end{verbatim}

Figure \ref{fig:pop-states} has some interesting things in it:

\begin{itemize}
\item
  for the deterministic IPM (and the kernel-resampled IPM?) there are slightly more of the smallest plants and the largest plants in CF compared to FF (i.e.~more medium sized plants in FF).
\item
  For the kernel-resampled IPM (random effect of year), the fluctuations are extremely similar between CF and FF
\item
  For the parameter-resampled IPM (DLNM) the size structure of the population is a LOT more variable in FF.
  This makes sense as we know lagged effects are more important in fragments.
\item
  Also, the fluctuations in size structure in CF do not match the fluctuations in FF as well (can see this by the increased spread of points in Figure \ref{fig:pop-states}B)
\item
  Also, in the parameter-resampled IPM (and only in this one), we see a shift toward smaller plants in FF compared to CF
\end{itemize}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

@@

\hypertarget{credit-statement}{%
\section{CRediT Statement}\label{credit-statement}}

@@

\hypertarget{data-availability-statement}{%
\section{Data Availability Statement}\label{data-availability-statement}}

@@

\hypertarget{tables}{%
\section{Tables}\label{tables}}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3333}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}@{}}
\caption{\label{tab:aic} Comparison of vital rate models used to build IPM. The `Effect of Environment' column describes how environmental effects were included in models. Those with `none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. `edf' is the estimated degrees of freedom of the penalized GAM. ∆AIC is calculated within each habitat and vital rate combination. ∆AIC within 2 indicates models are equivalent.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Vital Rate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Effect of Environment
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
∆AIC
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Vital Rate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Effect of Environment
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
∆AIC
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
CF & Survival & Random effect of year & 43.26 & 0 \\
CF & Survival & DLNM & 19.72 & 78.92 \\
CF & Survival & None & 4.976 & 260 \\
CF & Growth & Random effect of year & 78.43 & 0 \\
CF & Growth & DLNM & 23.87 & 158.5 \\
CF & Growth & None & 7.81 & 1896 \\
CF & Flowering & DLNM & 19.59 & 0 \\
CF & Flowering & Random effect of year & 17.19 & 1.627 \\
CF & Flowering & None & 7.468 & 381.9 \\
CF & Seedling survival & None & 1 & 0 \\
CF & Seedling survival & Random effect of year & 1.817 & 1.386 \\
CF & Seedling survival & DLNM & 4.008 & 1.528 \\
CF & Seedling growth & Random effect of year & 9.475 & 0 \\
CF & Seedling growth & DLNM & 8.952 & 2.902 \\
CF & Seedling growth & None & 1 & 172.3 \\
FF & Survival & DLNM & 14.95 & 0 \\
FF & Survival & Random effect of year & 19.21 & 35.68 \\
FF & Survival & None & 4.333 & 51.25 \\
FF & Growth & DLNM & 25.18 & 0 \\
FF & Growth & Random effect of year & 37.84 & 200 \\
FF & Growth & None & 5.599 & 382.8 \\
FF & Flowering & DLNM & 20.61 & 0 \\
FF & Flowering & Random effect of year & 13.81 & 27.4 \\
FF & Flowering & None & 5.007 & 101.7 \\
FF & Seedling survival & DLNM & 5.574 & 0 \\
FF & Seedling survival & Random effect of year & 5.088 & 5.721 \\
FF & Seedling survival & None & 1 & 6.491 \\
FF & Seedling growth & Random effect of year & 6.25 & 0 \\
FF & Seedling growth & DLNM & 8.182 & 2.29 \\
FF & Seedling growth & None & 1 & 5.745 \\
\end{longtable}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=1in,left=1in,bottom=1in,right=1in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\normalsize}
\usepackage{lineno}\linenumbers
\usepackage{sectsty}\sectionfont{\centering}
\usepackage{sectsty}\subsectionfont{\centering}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Context-dependent consequences of including lagged effects in demographic models},
  pdfkeywords={integral projection models, environmental
stochasticity, lagged effects},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Context-dependent consequences of including lagged effects in
demographic models}

\author[1]{Eric R. Scott}
\author[1,2,3]{Emilio M. Bruna}

\affil[1]{Department of Wildlife Ecology and Conservation, University of
Florida, Gainesville, Florida 32611-0430}
\affil[2]{Center for Latin American Studies, University of Florida,
Gainesville, Florida 32611-5530}
\affil[3]{Biological Dynamics of Forest Fragments Project, INPA-PDBFF,
CP 478, Manaus, Amazonas 69011-970 Brazil}

\date{(draft: 20 July 2023)}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, breakable]}{\end{tcolorbox}}\fi

@@

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\hypertarget{paragraph-1}{%
\subsubsection{\texorpdfstring{\textbf{Paragraph
1}}{Paragraph 1}}\label{paragraph-1}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Structured demographic models, such as Integral Projection Models
  (IPMs) and matrix models, have emerged as invaluable tools in ecology,
  evolution, and conservation biology.
\end{enumerate}

They provide insights into \_\_\_\_, the evolution of life-history
strategies, and life-history stages and ecological processes at which
conservation intervention is likely to have the greatest impact on
population growth rate.

Their flexibility hasmade them increasingly applicable to species with
differtent life histories.

however, there are still a number of challenges to address, namely
\_\_\_\_ , \_\_\_\_, and \_\_\_\_.

\hypertarget{paragraph-2}{%
\subsubsection{\texorpdfstring{\textbf{Paragraph
2}}{Paragraph 2}}\label{paragraph-2}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  It has long been recognized that there is the potential for lagged
  effects.
\item
  Including lagged effects in models has been a major technical
  challenge, but there are now multiple approaches for doing so.
\item
  The studies assessing the potential for lagged effects on vital rates
  find that they indeed appear to be prevalent (Evers et al. 2021; Scott
  et al. 2022).
\end{enumerate}

\hypertarget{paragraph-3}{%
\subsubsection{\texorpdfstring{\textbf{Paragraph
3}}{Paragraph 3}}\label{paragraph-3}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Lagged effects could be different in different habitat types.
\item
  This could be a big reason for differences between habitats.
\end{enumerate}

\hypertarget{paragraph-4}{%
\subsubsection{\texorpdfstring{\textbf{Paragraph
4}}{Paragraph 4}}\label{paragraph-4}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Here we\ldots{}
\item
  We conducted these analyses with both deterministic and stochastic
  IPMs.
\end{enumerate}

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

\hypertarget{demographic-methods-and-data}{%
\subsubsection{\texorpdfstring{\emph{Demographic methods and
data}}{Demographic methods and data}}\label{demographic-methods-and-data}}

Overview of the heliconia project and data

\hypertarget{construction-of-integral-projection-models}{%
\subsubsection{\texorpdfstring{\emph{Construction of Integral Projection
Models}}{Construction of Integral Projection Models}}\label{construction-of-integral-projection-models}}

@@

We built three classes of IPMs for comparison which each required
different functional forms of their underlying vital rates models. The
simplest IPM was a general, density-independent, deterministic IPM with
four sub-kernels: growth and survival (\(P\), (\textbf{eq:P?})),
fecundity (\(F\), i.e.~production of new seedlings, (\textbf{eq:F?})),
probability of staying a seedling (always 0), and recruitment (\(R\),
i.e.~seedling survival and establishment, (\textbf{eq:R?}))
((\textbf{eq:mature?}; \textbf{eq:sdlg?}), Figure @ref(fig:lifecycle)).
The probability of staying a seedling, was always equal to zero, since
our definition of seedlings was first year plants only.

@@

To build the general, density-independent, stochastic, kernel-resampled
IPMs, we included environmental stochasticity in all vital rate models
built using the long term demographic datset by adding a random effect
of year (Figure @ref(fig:lifecycle)). The random effect of year was
included using a factor--smooth interaction which allowed the
relationship between plant size and vital rates to vary in functional
form among transition years. The kernel-resampling approach is to
generate kernels corresponding to each transition year in the
demographic dataset using the random smooths for year, and to iterate
the IPM by drawing from these randomly. This is equivalent to the matrix
selection approach for matrix population models described by Caswell
(2001).

For the third method, we modeled the impacts of drought on vital rates
explicitly and created general, density-independent, stochastic,
parameter-resampled IPMs (\emph{sensu} Metcalf et al. (2015)). We
calculated the standardized precipitation evapotranspiraton index (SPEI)
for our site using a published gridded dataset based on ground
measurements (Xavier et al. 2016) as described in Scott et al. (2022).
For all vital rate models fit using the long term demographic dataset,
we modeled delayed effects of SPEI using distributed lag non-linear
models with a maximum lag of 36 months(Scott et al. 2022) (Figure
@ref(fig:lifecycle)). To iterate these parameter-resampled IPMs, a
random sequence of SPEI values was created by sampling years of the
observed monthly SPEI data. Then, 36 month lags are calculated for each
year starting in February (the month of the demographic census). These
values are then used to predict fitted values from the vital rates
models, generating different kernels at each iteration of the IPM. With
this method, the kernels of successive iterations are not entirely
independent because the SPEI values used in calculating vital rates
include values used in the previous two iterations, but they are
ergodic.

@@

\hypertarget{statistical-analyses}{%
\subsubsection{\texorpdfstring{\emph{Statistical
analyses}}{Statistical analyses}}\label{statistical-analyses}}

All about the stats.

\hypertarget{results}{%
\subsection{Results}\label{results}}

For all vital rates estimated using the long term demographic dataset,
the DLNM model fit the best (dAIC = 0) followed by the model with a
random effect of year, followed by the deterministic model (Table
@ref(tab:aic)).\\

Population growth rates were consistently higher in continuous forest
compared to forest fragments across IPM types (Table @ref(tab:lambdas)).

The time to iterate the DLNM models is much higher than than
deterministic and kernel-resampled.

The greater use of computational resources is likely a result of
\texttt{predict()} being much slower for GAMs with 2D smooths because
the number of knots is much higher compared to the GAMs used for the
vital rates models in the determinsitic and kernel-resampled IPMs.

Figure @ref(fig:pop-states) has some interesting things in it:

\begin{itemize}
\item
  for the deterministic IPM (and the kernel-resampled IPM?) there are
  slightly more of the smallest plants and the largest plants in CF
  compared to FF (i.e.~more medium sized plants in FF).
\item
  For the kernel-resampled IPM (random effect of year), the fluctuations
  are extremely similar between CF and FF
\item
  For the parameter-resampled IPM (DLNM) the size structure of the
  population is a LOT more variable in FF. This makes sense as we know
  lagged effects are more important in fragments.
\item
  Also, the fluctuations in size structure in CF do not match the
  fluctuations in FF as well (can see this by the increased spread of
  points in Figure @ref(fig:pop-states)B)
\item
  Also, in the parameter-resampled IPM (and only in this one), we see a
  shift toward smaller plants in FF compared to CF
\end{itemize}

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

@@

\hypertarget{acknowledgments}{%
\subsection{Acknowledgments}\label{acknowledgments}}

@@

\hypertarget{credit-statement}{%
\subsection{CRediT Statement}\label{credit-statement}}

@@

\hypertarget{data-availability-statement}{%
\subsection{Data Availability
Statement}\label{data-availability-statement}}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1389}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3333}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1111}}@{}}
\caption{(\#tab:aic) Comparison of vital rate models used to build IPM.
The `Effect of Environment' column describes how environmental effects
were included in models. Those with `none' were used to build
deterministic IPMs; those with a random effect of year were used to
build stochastic, kernel-resampled IPMs; and those with a distributed
lag non-linear model (DLNM) were used to build stochastic,
parameter-resampled IPMs. `edf' is the estimated degrees of freedom of
the penalized GAM. ΔAIC is calculated within each habitat and vital rate
combination. ΔAIC within 2 indicates models are
equivalent.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Vital Rate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Effect of Environment
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
ΔAIC
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Habitat
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Vital Rate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Effect of Environment
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
edf
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
ΔAIC
\end{minipage} \\
\midrule()
\endhead
CF & Survival & Random effect of year & 43.26 & 0 \\
CF & Survival & DLNM & 19.72 & 78.92 \\
CF & Survival & None & 4.976 & 260 \\
CF & Growth & Random effect of year & 78.43 & 0 \\
CF & Growth & DLNM & 23.87 & 158.5 \\
CF & Growth & None & 7.81 & 1896 \\
CF & Flowering & DLNM & 19.59 & 0 \\
CF & Flowering & Random effect of year & 17.19 & 1.627 \\
CF & Flowering & None & 7.468 & 381.9 \\
CF & Seedling survival & None & 1 & 0 \\
CF & Seedling survival & Random effect of year & 1.817 & 1.386 \\
CF & Seedling survival & DLNM & 4.008 & 1.528 \\
CF & Seedling growth & Random effect of year & 9.475 & 0 \\
CF & Seedling growth & DLNM & 8.952 & 2.902 \\
CF & Seedling growth & None & 1 & 172.3 \\
FF & Survival & DLNM & 14.95 & 0 \\
FF & Survival & Random effect of year & 19.21 & 35.68 \\
FF & Survival & None & 4.333 & 51.25 \\
FF & Growth & DLNM & 25.18 & 0 \\
FF & Growth & Random effect of year & 37.84 & 200 \\
FF & Growth & None & 5.599 & 382.8 \\
FF & Flowering & DLNM & 20.61 & 0 \\
FF & Flowering & Random effect of year & 13.81 & 27.4 \\
FF & Flowering & None & 5.007 & 101.7 \\
FF & Seedling survival & DLNM & 5.574 & 0 \\
FF & Seedling survival & Random effect of year & 5.088 & 5.721 \\
FF & Seedling survival & None & 1 & 6.491 \\
FF & Seedling growth & Random effect of year & 6.25 & 0 \\
FF & Seedling growth & DLNM & 8.182 & 2.29 \\
FF & Seedling growth & None & 1 & 5.745 \\
\bottomrule()
\end{longtable}

\newpage

\begin{table}

\caption{table caption}
\centering
\begin{tabular}[t]{lc}
\toprule
IPM Type & mean time (min.)\\
\midrule
Deterministic & 0.02\\
Stochastic, kernel-resampled & 0.07\\
Stochastic, parameter-resampled & 87.12\\
\bottomrule
\end{tabular}
\end{table}

\newpage
\newpage
